\hspace{1cm}Semantic interference during encoding and retrieval \newline Evidence from self-paced reading and event-related potentials

### 2023-10-08 paper/main.tex
edited up to plausibility rating

Previous work has shown clear evidence for this kind of retrieval interference \parencite{vandyke07, vandyke_mcelree06, vandyke_mcelree2011, nicenboim,  vandyke_lewis03,jaeger_etal_2017}.
However, two important questions remain unanswered. First, as discussed in \cite{jaeger_etal_2017}, almost all the published studies on interference effects are severely underpowered. Low power has the consequence that published estimates will tend to be overestimates or misestimates of the effect size \parencite{vasishth2018_signficancefilter}. Over-/misestimates have important implications for theory evaluation. One example is discussed in \cite{vasishth2018_signficancefilter}, where a key prediction of surprisal theory \parencite{levy&keller_2013} could not be validated when a larger-sample study was conducted; by contrast, the original small-sample study  \parencite{levy&keller_2013} showed suggestive evidence consistent with the theory. Another example is the recent heated debate on prediction in sentence comprehension, which was largely focused on the results of low-powered studies, leading to potentially misleading conclusions \parencite{nicenboim_etal_2020}.
Second, although it is clear and uncontroversial  \parencite[e.g.,][]{vandyke07,mertzen} that syntactic and semantic cues are involved in sentence processing, it is not obvious which cues are relevant for a particular retrieval event. For example, \cite{dillon2013} argued, following \cite{Sturt2003}, that the gender cue was either downweighted or not used at all when an antecedent-reflexive dependency needs to be built; instead, the antecedent of a reflexive could be identified using the syntactic principle A of the binding theory, which correctly identifies the target for retrieval.  \cite{dillon2013} suggested that antecedent-reflexive dependencies were immune to interference \parencite[cf. ][]{jaeger_etal_2020,yadav2021individual}. The broader implication here is that the cues used during retrieval can modulate processing difficulty in possibly surprising ways.

@@

We aimed to investigate the role of syntactic and semantic cues during sentence comprehension with a series of experiments using self-paced reading and EEG. All experiments used the 2 $\times$ 2 design by \citet{vandyke07} crossing syntactic and semantic interference. This is -- to our knowledge -- the first ERP study on this classic design investigating interference during subject-verb dependency resolution. The SPR experiments were conducted web-based via Prolific and the EEG experiments required participants to come into the lab. All experiments were run with native speakers of German. Because the high signal-to-noise ratio of EEG data requires a large number of trials per condition and participant, we constructed 120 items which were used in the EEG and SPR experiments. It would have been very taxing for the participants to read all 120 items in one experimental session, therefore we decided to split the experiment into two sessions. The ERP data that we present here, comes only from participants who completed both sessions. In contrast, most of the SPR participants (570 out of 774) finished only one session. This was intentional because we found that participants' reading speed in the second session was increased, probably due to increased familiarity with the task. Figure \ref{fig:project_str} shows the structure of the present study, i.e., how many participants saw which subset of the items and whether these participants completed one or two experimental sessions.

\begin{figure}[H]
    \caption{Structure of the present study. In total, 120 items were used for the SPR (red branches) and the ERP experiments (blue branches). The color brightness reflects how many sessions were completed by each participant (lighter color: two sessions, darker color: one session).}
    \label{fig:project_str}
    \centering
    \includegraphics[width=0.85\textwidth]{images/pandora_project_structure_figure.pdf}
\end{figure}

\subsection{Materials}
As the same materials were used for all our experiments, we describe them here once before we describe the specifics of the experiments, separately. The materials for the study consisted of 120 items with 4 conditions in a Latin square design shown in \ref{ex:materials}. The items were constructed following \citeauthor{vandyke07}'s (\citeyear{vandyke07}) items and partially re-used  \citeauthor{mertzen}'s (\citeyear{mertzen}) items.

@@

Comprehension of the sentences in \ref{ex:materials} requires a memory retrieval operation at the verb \textit{trank} (`drank') to retrieve its non-adjacent subject \textit{Witwer} (`widower'). During this retrieval, the distractor \textit{Einbrecher} (`burglar') / \textit{Verlust} (`loss') might cause interference and impede comprehension. Syntactic interference is manipulated via the syntactic status of the distractor, which is either subject (\ref{ex:hisyn}) or part of a prepositional phrase (\ref{ex:losyn}). During the retrieval of the subject, a distractor in subject position is assumed to cause more syntactic interference than a distractor inside a prepositional phrase. Semantic interference is manipulated via the animacy of the distractor. During the retrieval of the subject of the verb \textit{trank} (`drank'), the animate distractor  \textit{Einbrecher} (`burglar') is assumed to cause high semantic interference, while the inanimate distractor \textit{Verlust} (`loss') is assumed to cause low semantic interference. In contrast to the previous studies (see Introduction), we inserted a second pre-critical adverb, i.e., a pre-pre-critical region, to absorb potential spill-over effects from the embedded clause(s). So, potential wrap-up effects from the embedded clause(s) could spill-over into the pre-pre-critical region and the pre-critical region would be further away from clause boundaries. This way potential pre-critical effects are less likely to be caused by structural confounds.

\subsubsection{Plausibility norming}
\citeauthor{vandyke07} conducted a pre-test to ensure that the distractor in high semantic interference conditions was a sufficiently probable subject of the critical verb, i.e., that it could potentially induce semantic interference. For the same purpose, we conducted a plausibility norming experiment.

\paragraph{Participants}
Forty participants (18 female, 22 male, mean age: 27.3 years old, age range: 21 - 39 years old) from the SPR experiments were re-invited for the plausibility norming one month after they completed the SPR experiment.

\paragraph{Materials}
For each item quadruplet, four simple SVO sentences were constructed, so that each of the nouns in the item once functioned as the subject of the critical verb. The norming materials corresponding to the item despicted in \ref{ex:materials} are shown in \ref{ex:materials_norming}.

@@

\paragraph{Procedure}
The plausibility norming experiment was implemented in PCIbex \citep{pcibex} and was conducted online. Participants were asked to judge the plausibility of the sentences on a 7-point scale ranging from absolute implausible (1) to absolute plausible (7). All sentences belonging to one item were shown on one screen, so that they could be judged relative to each other. The plausibility norming experiment included 120 sentences from a different experiment which functioned as fillers. The participants received \pounds 3.40 as compensation for the norming experiment. A demonstration of the plausibility norming experiment can be accessed via \hyperlink{https://farm.pcibex.net/r/WIeKYM/}{https://farm.pcibex.net/r/WIeKYM/}.

\paragraph{Results}
The plausibility ratings showed that the semantic manipulation worked as intended. The subject received the highest rating (mean: 6.5, sd: 0.9) and the animate distractor (high semantic interference conditions) was almost as plausible (mean: 6.1, sd: 1.3). In contrast, the inanimate distractor (low semantic interference conditions) was very implausible (mean: 1.7, sd: 1.4). The introduction noun which was not manipulated was tested as well and it received a high plausibility rating (mean: 6.0, sd: 1.4). These results are illustrated in Figure \ref{fig:plausibility_rating}. The plausibility ratings are similar to those reported for \citeauthor{vandyke07}'s (\citeyear{vandyke07}) Experiment 3.

>>>>

Previous work has shown clear evidence for this kind of retrieval interference \parencite{vandyke07, vandyke_mcelree06, vandyke_mcelree2011, nicenboim,  vandyke_lewis03,jaeger_etal_2017}.
However, two important questions remain unanswered. First, as discussed in \cite{jaeger_etal_2017}, almost all the published studies on interference effects are severely underpowered. Low power has the consequence that published estimates will tend to be overestimates or misestimates of the effect size \parencite{vasishth2018_signficancefilter}. Over-/misestimates have important implications for theory evaluation. One example is discussed in \cite{vasishth2018_signficancefilter}, where a key prediction of surprisal theory \parencite{levy&keller_2013} could not be validated when a larger-sample study was conducted; by contrast, the original small-sample study  \parencite{levy&keller_2013} showed suggestive evidence consistent with the theory. Another example is the recent heated debate on prediction in sentence comprehension, which was largely focused on the results of low-powered studies, leading to potentially misleading conclusions \parencite{nicenboim_etal_2020}.
Second, although it is clear and uncontroversial  \parencite[e.g.,][]{vandyke07,mertzen} that syntactic and semantic cues are involved in sentence processing, it is not obvious which cues are relevant for a particular retrieval event. For example, \cite{dillon2013} argued, following \cite{Sturt2003}, that the gender cue was either downweighted or not used at all when an antecedent-reflexive dependency needs to be built; instead, the antecedent of a reflexive could be identified using the syntactic principle A of the binding theory, which correctly identifies the target for retrieval.  Using computational modeling, \cite{dillon2013} suggested that antecedent-reflexive dependencies were immune to interference, and that the reason for this is that principle A of the binding theory is the primary driver of the cue-based retrieval process \parencite[cf. ][]{jaeger_etal_2020,yadav2021individual}. This was a surprising discovery because a priori, cue-based retrieval accounts do not allow for exceptions to retrieval interference in particular syntactic dependency types.
More broadly, an important implication of the work initiated by \cite{dillon2013}, among others, is that  the precise details of which cues are used during retrieval can affect processing difficulty in possibly surprising ways.

@@

Reading studies have been the dominant method used to investigate retrieval interference \parencite[see the review in][]{jaeger_etal_2017}.  We aimed to investigate the role of syntactic and semantic cues during sentence comprehension with higher-powered experiments using  self-paced reading as well as EEG, which to our knowledge has never been used to investigate interference during subject-verb dependency resolution

All experiments used the 2 $\times$ 2 design developed by \citet{vandyke07}; the design crosses syntactic and semantic interference.

% Wrong location for all this detail:
%This is -- to our knowledge -- the first . The SPR experiments were conducted web-based via Prolific and the EEG experiments required participants to come into the lab. All experiments were run with native speakers of German. Because the high signal-to-noise ratio of EEG data requires a large number of trials per condition and participant, we constructed 120 items which were used in the EEG and SPR experiments. It would have been very taxing for the participants to read all 120 items in one experimental session, therefore we decided to split the experiment into two sessions. The ERP data that we present here, comes only from participants who completed both sessions. In contrast, most of the SPR participants (570 out of 774) finished only one session. This was intentional because we found that participants' reading speed in the second session was increased, probably due to increased familiarity with the task. Figure \ref{fig:project_str} shows the structure of the present study, i.e., how many participants saw which subset of the items and whether these participants completed one or two experimental sessions.

\begin{figure}[H]
    \caption{The structure of the present study. In total, 120 items were used for the SPR (red branches) and the ERP experiments (blue branches). The color brightness reflects how many sessions were completed by each participant (lighter color: two sessions, darker color: one session).}
    \label{fig:project_str}
    \centering
    \includegraphics[width=0.85\textwidth]{images/pandora_project_structure_figure.pdf}
\end{figure}

\subsection{Materials}

As the same materials were used for all our experiments, we describe them here once before we describe the specifics of the experiments, separately. The studies had 120 items with four conditions in a Latin square design shown in \ref{ex:materials}. The items were constructed following \citeauthor{vandyke07}'s (\citeyear{vandyke07}) items and partially re-used  \citeauthor{mertzen}'s (\citeyear{mertzen}) items.

@@

Comprehension of the sentences in \ref{ex:materials} requires a memory retrieval operation at the verb \textit{trank} (`drank') to retrieve its non-adjacent subject \textit{Witwer} (`widower'). During this retrieval, the distractor \textit{Einbrecher} (`burglar') / \textit{Verlust} (`loss') might cause interference and impede comprehension.

Syntactic interference is manipulated via the syntactic status of the distractor, which is either subject (\ref{ex:hisyn}) of an embedded clause, or part of a prepositional phrase (\ref{ex:losyn}). Following \cite{vandyke07}, the assumption here is that  during the retrieval of the subject, a distractor in any subject position causes more syntactic interference than a distractor inside a prepositional phrase.

Semantic interference is manipulated via the animacy of the distractor. During the retrieval of the subject of the verb \textit{trank} (`drank'), the animate distractor  \textit{Einbrecher} (`burglar') is assumed to cause high semantic interference, while the inanimate distractor \textit{Verlust} (`loss') is assumed to cause low semantic interference.

It is well-known \parencite[e.g., ][]{rayner2000effect} that the last word in a clause-final position causes a slowdown in reading; such a slowdown has the potential to spill over to the subsequent regions, possibly affecting the estimated effects at the pre-critical and critical regions. The spillover of longer reading times can be problematic because the standard deviation increases as well \parencite{wagenmakers2007linear}; such an increase in standard deviation would drastically lower statistical power. In order to attenuate spillover from the clause-final word in the embedded clause preceding the critical region, we inserted two adverbs. This way, the critical region will be less affected by the clause-boundary slowdown. This design also has the advantage that any potential effects seen at the pre-critical region (the adverb preceding the verb) are less likely to be due to spillover; it is important to look at the pre-critical region as previous work with this design found effects at the pre-critical region (we revisit this detail in the General discussion).

\subsubsection{Plausibility norming}
\citeauthor{vandyke07} conducted a pre-test to ensure that the distractor in high semantic interference conditions was a sufficiently probable subject of the critical verb, i.e., that it could potentially induce semantic interference. Following \citeauthor{vandyke07}, we also conducted a plausibility norming experiment.

\paragraph{Participants}
Forty participants (18 female, 22 male, mean age: 27.3 years old, age range: 21 - 39 years old) from the SPR experiments were re-invited for the plausibility norming one month after they completed the SPR experiment. We used the same participants who participated in the SPR study in order to ensure that the ratings for the norming came from the same pool of participants that did the reading study.

\paragraph{Materials}
For each item quadruplet, four simple sentences with subject-verb-object word order were constructed, so that each of the nouns in the item once functioned as the subject of the critical verb. The norming materials corresponding to the item depicted in \ref{ex:materials} are shown in \ref{ex:materials_norming}.

@@

\paragraph{Procedure}
The plausibility norming experiment was implemented in PCIbex \citep{pcibex} and was conducted online. Participants were asked to judge the plausibility of the sentences on a 7-point scale ranging from absolutely implausible (1) to absolutely plausible (7). All sentences belonging to one item were shown on one screen, so that they could be judged relative to each other. The plausibility norming experiment included 120 sentences from a different experiment which functioned as fillers. The participants received \pounds 3.40 as compensation for the norming experiment. A demonstration of the plausibility norming experiment can be accessed via \hyperlink{https://farm.pcibex.net/r/WIeKYM/}{https://farm.pcibex.net/r/WIeKYM/}.

\paragraph{Results}
The plausibility ratings showed that the semantic manipulation worked as intended. The subject received the highest rating (mean: 6.5, sd: 0.9) and the animate distractor (high semantic interference conditions) was almost as plausible (mean: 6.1, sd: 1.3). In contrast, the inanimate distractor (low semantic interference conditions) was very implausible (mean: 1.7, sd: 1.4). The first noun in the target sentences, which was not part of the experimental manipulation, was tested as well and it received a high plausibility rating (mean: 6.0, sd: 1.4). These results are shown in Figure \ref{fig:plausibility_rating}. The plausibility ratings are similar to those reported for \citeauthor{vandyke07}'s (\citeyear{vandyke07}) Experiment 3.
